---
title: "MLB's Sticky Stuff Scandal"
author: "Abigail Smith"
format: revealjs
---

```{r message=FALSE, warning=FALSE, echo = FALSE}
library(here)
library(readr)
library(tidyverse)
library(plotly)

# Reading in the pitching data set
pitching_df <- read_csv(here("data/pitchingstats.csv"))

# Reading in each individual attendance data set
attendancefull <- read_csv(here("data/attendancefull.csv"))

# Reading in the world series data set
worldseriesviewing <- read_csv(here("data/world-series-ratings.csv"))

# Cleaning pitching
pitching_df <- pitching_df |> rename(pitcher = `last_name, first_name`)

pitching_cleaner <- pitching_df |> pivot_longer(c("sl_avg_speed", "ch_avg_speed", "cu_avg_speed", "si_avg_speed"), names_to = "speedpitch_type", values_to =  "pitch_speed") |> pivot_longer(c("sl_avg_spin", "ch_avg_spin", "cu_avg_spin", "si_avg_spin"), names_to = "spinpitch_type", values_to =  "spin_rate") |>separate(speedpitch_type, into = c("pitchtype", "speed"), sep = "_avg_") |> separate(spinpitch_type, into = c("pitch_type", "spin"), sep = "_avg_") |> select(!(c(pitchtype, speed, spin))) 

pitching_cleaner <- pitching_cleaner |> mutate(pitch_type = case_when(pitch_type == "sl" ~ "slider", pitch_type == "ch" ~ "changeup", pitch_type == "cu" ~ "curveball", pitch_type == "si" ~ "sinker")) 

# Cleaning up attendance and tv viewership dfs
attendance_smaller <- attendancefull |> select(Tm, Attendance, `Attend/G`, Time, `Est. Payroll`, year)

tv_views_smaller <- worldseriesviewing |> select(c(year, average_audience)) |> filter(year >= 2018 & year <= 2023)
```

## Pitching Data

-   pitching_cleaner
    -   2018-2024
    -   768 pitchers
    -   6 dependent variables (HR, K%, BB%, HBP, xBA, xSLG)
    -   2 independent variables (spin rate and pitch type (curveball, changeup, slider, sinker))
    
# Attendance 
-   attendance_smaller
    -   2018-2024
    -   30 teams
    -   3 variables (attendance, attendance per game, year)

# TV Viewership Data 
-   tv_views_smaller
    -   2018-2022
    -   1 variable (network)

## Question of interest

Using spin rate to inspect how big of an effect did the sticky stuff scandal had on baseball as a whole.

## Pitching scatterplot

```{r}
ggplot(data = pitching_cleaner, aes(x= spin_rate, y = home_run, colour = pitch_type)) + geom_point() + theme_minimal() + facet_wrap(~year) + labs(title ="faceted scatterplot of spinrate against home runs")
```

# Pitching histogram

```{r}
ggplot(data = pitching_cleaner, aes(x = spin_rate)) + geom_histogram() + facet_wrap(~year) + theme_minimal() + labs(title = "Histogram of spin rate facted by year")
```

## Attendance lineplot

```{r}
ggplot(data = attendance_smaller, aes(x = year, y = `Attend/G`, colour = Tm)) + geom_line() + theme_minimal() + labs(title = "Average Attendance per Game Over Time ")
```

# TV viewership barplot

```{r}
ggplot(data = tv_views_smaller, aes(x = year, y = average_audience)) + geom_col() + theme_minimal() + labs(title = "Average TV Audience per World Series Game In Each Year ")
```
