---
title: "Final Project"
format: html
---

# Abstract

My project is focused on the "Sticky Stuff" scandal in Major League Baseball (MLB) in 2021. The scandal involved MLB pitchers using sticky substances on their hands to increase their spin rate and decrease pitch speed. While pitchers have always used simple substances such as sunscreen and rosin to gain grip, pitchers were now using materials such as Spider Tack an extremely high stick glue like substance to gain grip. Furthermore, some teams had even begun hiring chemists to develop specific substances of extreme strength. Pitchers would hide this material on their glove or hat and then rub it on their hands to gain grip. The higher spin rates resulted in an increase in strike outs as it is harder for the batter to hit the ball when it has so much movement. Additionally, more batters were getting hit since pitchers had less control and the amount of home runs hit went down dramatically since again the ball grew harder to hit. These excessive changes in hitting patterns led to the MLB leading a major cut down on the substances, banning them altogether and integrating regular checks at the end of each inning to make sure pitchers are not using the materials. The MLB provides access to their StatCast data which is collected through several tracking cameras dispersed throughout the field. This data comprises of relevant statistics such as spin rate in addition to batting statistics which can show the overall trends over time.

# Introduction

For the project I am interested in investigating the impact of the sticky stuff scandal on baseball as a whole. I will do so by creating a Shiny App using the StatCast data as well as some attendance and TV viewership data I found online. In looking at the attendance data I am curious to see if there is a noticeable pattern between the trend in attendance and the trends in spin rate. More specifically, I would like to see if attendance increased after the MLB cracked down on sticky substances. My theory is that I will see an increase in attendance following 2021, as generally speaking attendance is higher when the amount of home runs is higher. Take for example, the steroid era in the early 2000s, home runs were at an all time high because so many players were taking steroids, and attendance was extremely high as well. This leads me to believe that something like the sticky stuff scandal would have the opposite effect and harm attendance. I also think a brief glance at some world series tv viewership data I found will be relevant to this same idea. Although in this situation the sample size is smaller, I still think that a trend in world series viewership reflects on the viewership and popularity of baseball in that season as a whole. I would also like to see just how evident the impact of spin rate on batting statistics such as home runs, batting average, and slugging average are. Similarly, I want to see how spin rate impacted other pitching statistics, such as strike out percentage and the number of players hit by pitch. On a simple note, I would like to take a glance at pitching how spin rate has changed over time just to see how evident the direct influence of the sticky stuff scandal is.

# Pitching dataset description

The initial data set that I read in is from MLB's Statcast Feature, it is accessible on the [Baseball Savant website](https://baseballsavant.mlb.com/leaderboard/custom?year=2024%2C2023%2C2022%2C2021%2C2020%2C2019%2C2018&type=pitcher&filter=&min=q&selections=pa%2Chome_run%2Ck_percent%2Cbb_percent%2Cxba%2Cxslg%2Csl_avg_speed%2Csl_avg_spin%2Cch_avg_speed%2Cch_avg_spin%2Ccu_avg_speed%2Ccu_avg_spin%2Csi_avg_speed%2Csi_avg_spin&chart=false&x=pa&y=pa&r=no&chartType=beeswarm&sort=pa&sortDir=desc). I adjusted the data frame to include pitching statistics for individual pitchers in the years 2018-2024 for the whole MLB. The below table describes each of the variables in the data set.

| Variable         | Description                                                                                     |
|----------------|--------------------------------------------------------|
| `pitcher`        | The last and first name of the player.                                                          |
| `year`           | The year for the case.                                                                          |
| `home_run`       | The number of home runs allowed by the pitcher in that season.                                  |
| `k_percent`      | The strike out percentage, the number of strike outs per batter faced by the pitcher.           |
| `bb_percent`     | The walk percentage, the number of walks per batter faced by the pitcher.                       |
| `p_hit_by_pitch` | The number of batters hit by the pitcher in that season.                                        |
| `xba`            | Opposing players' batting average against the pitcher in that season.                           |
| `xslg`           | Opposing players' slugging average against the pitcher in that season.                          |
| `spin_rate`      | The spin rate for each individual pitch, measured in rotations per minute.                      |
| `pitch_type`     | The type of pitch each pitch is, which could be (curveball, sinker, slider, and changeup)       |

# Attendance datasets descriptions

Additionally, I used [basbeall-reference](https://www.baseball-reference.com/leagues/majors/2024-misc.shtml) for data on baseball attendance, in the years 2018-2024. For this data, it was all in individual data sets which I had to join together into one big one. In each dataset the variables are the same which I created a table for.

| Variable       | Description                                   |
|----------------|-----------------------------------------------|
| `Tm`           | The team's name.                              |
| `Attendance`   | Total attendance for home games that season.  |
| `Attendance/G` | Average attendance per home game that season. |
| `Year`         | The year that the case is from.               |

# TV Viewership dataset description

As a tentative tv viewership dataset I found is this one from [kaggle](https://www.kaggle.com/datasets/mattop/world-series-baseball-television-ratings?resource=download) that covers just world series game viewership between 2022 and 2024. I am trying to find a better dataset that is general tv viewership, however it seems that most if not all avialable mlb viewership data is focused solely on the world series. However, conclusions can still be drawn from this data as the world series are the most watched games in the baseball season, and if viewership is down for those its definitely down for the entire season.

| Variable           | Description                                         |
|--------------------|-----------------------------------------------------|
| `year`             | The year of the world series.                       |
| `average_audience` | The average tv audience for the whole world series. |

# Packages

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(here)
library(readr)
library(tidyverse)
```

# Data

```{r}
# Reading in the pitching data set
pitching_df <- read_csv(here("data/pitchingstats.csv"))

# Reading in each individual attendance data set
attendance_2018 <- read_csv(here("data/2018attendance.csv"))
attendance_2019 <- read_csv(here("data/2019attendance.csv"))
attendance_2020 <- read_csv(here("data/2020attendance.csv"))
attendance_2021 <- read_csv(here("data/2021attendance.csv"))
attendance_2022 <- read_csv(here("data/2022attendance.csv"))
attendance_2023 <- read_csv(here("data/2023attendance.csv"))
attendance_2024 <- read_csv(here("data/2024attendance.csv"))

# Reading in the world series data set
worldseriesviewing <- read_csv(here("data/world-series-ratings.csv"))
```

```{r}
# Creating a year variable in each of the attendance data sets
attendance_2018$year <- 2018
attendance_2019$year <- 2019
attendance_2020$year <- 2020
attendance_2021$year <- 2021
attendance_2022$year <- 2022
attendance_2023$year <- 2023
attendance_2024$year <- 2024
```

```{r}
# Create 3 joined data sets of the attendance data sets
attendance2018_2019 <- full_join(attendance_2018, attendance_2019)
attendance2020_2021 <- full_join(attendance_2020, attendance_2021)
attendance2022_2023 <- full_join(attendance_2022, attendance_2023)

# Join those 3 data sets together into 2 datasets
attendance2018_2021 <- full_join(attendance2018_2019, attendance2020_2021)
attendance2022_2024 <- full_join(attendance2022_2023, attendance_2024)

# Join the last two big data sets in one big complete attendance data set
attendancefull <- full_join(attendance2018_2021, attendance2022_2024)
```

```{r}
# Save the attendance data set as a CSV so that I do not lose track of it
write_csv(attendancefull, "attendancefull.csv")
```

# Cleaning and tidying the pitching data

```{r}
pitching_df <- pitching_df |> rename(pitcher = `last_name, first_name`)
```

```{r}
pitching_cleaner <- pitching_df |> pivot_longer(c("sl_avg_spin", "ch_avg_spin", "cu_avg_spin", "si_avg_spin"), names_to = "pitch_type", values_to =  "spin_rate")

pitching_cleaner <- pitching_cleaner |> mutate(pitch_type = str_remove(pitch_type, "_avg_spin")) |>  mutate(pitch_type = case_when(pitch_type == "sl" ~ "slider", pitch_type == "ch" ~ "changeup", pitch_type == "cu" ~ "curveball", pitch_type == "si" ~ "sinker"))  

pitching_cleaner <- pitching_cleaner |> select(!c(sl_avg_speed, ch_avg_speed, si_avg_speed, cu_avg_speed, pa, player_id)) 
```

# Cleaning up attendance and tv viewership dataframes

I do not need all of these variables in the data frame so I am going to get rid of some of them

```{r}
attendance_smaller <- attendancefull |> select(Tm, Attendance, `Attend/G`, year)
```

Same deal here with the tv viewership

```{r}
tv_views_smaller <- worldseriesviewing |> select(c(year, average_audience)) |> filter(year >= 2018 & year <= 2023)
```

# Static visualizations for the shiny app

```{r}
attendance_smaller_reorder <- attendance_smaller |> mutate(Tm = fct_reorder(Tm, `Attend/G`))
ggplot(data = attendance_smaller_reorder, aes(x = Tm, y = `Attend/G`)) + geom_col() + coord_flip() + theme_minimal()
```

```{r}
ggplot(data = attendance_smaller, aes(x = year, y = `Attend/G`, colour = Tm)) + geom_line()
```

```{r}
ggplot(data = tv_views_smaller, aes(x = year, y = average_audience)) + geom_col()
```

```{r}
ggplot(data = pitching_cleaner, aes(x= spin_rate, y = home_run, colour = pitch_type)) + geom_point()
```

```{r}
ggplot(data = pitching_cleaner, aes(x = spin_rate)) + geom_histogram() + facet_wrap(~year) + theme_minimal() + labs(title = "Histogram of spin rate facted by year")
```

# Shiny app

```{r}
library(shiny)
year_choices_pitching <- pitching_cleaner |> distinct(year) |> pull(year)
y_var_choices <- names(pitching_cleaner)[c(3:8)]
pitch_choice <- pitching_cleaner |> distinct(pitch_type) |> pull(pitch_type)
x_var <- names(attendance_smaller)[c(2, 3)]
year_choices_attendance <- attendance_smaller|> distinct(year) |> pull(year)
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      radioButtons("year_sel", label = "Choose a year for pitching:", choices = year_choices_pitching),
      selectizeInput("pitch_sel", label = "Choose pitch type:",
                     choices = pitching_cleaner$pitch_type, selected = "slider",
                     multiple = TRUE),
      selectInput("y_var_sel", label = "Choose a response variable for pitching:", choices = y_var_choices),
      radioButtons("x_sel", label = "Choose an attendance variable:", choices = x_var),
      radioButtons("year_sel_attendance", label = "Choose a year for the attendance:", choices = year_choices_pitching),
      selectizeInput("tm_sel", label = "Choose a team:", choices = attendance_smaller$Tm, selected = "Arizona Diamondbacks", multiple = TRUE)
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Pitching", plotOutput("scatterplot"), plotOutput("spinhist"),  dataTableOutput("pitching_raw_data")),
        tabPanel("Attendance", plotOutput("attendance_line"), plotOutput("barplot"), dataTableOutput("attendance_raw_data")),
        tabPanel("TV Viewership", plotOutput("audience_bar"), dataTableOutput("audience_raw_data"))
      )
    )
  )
  
)


server <- function(input, output, session) {
 pitching_reactive <- reactive({
   pitching1 <- pitching_cleaner |> filter(year == input$year_sel) |> filter(pitch_type %in% input$pitch_sel)
})
 
output$scatterplot <- renderPlot({ggplot(pitching_reactive(), aes(x= spin_rate, y = .data[[input$y_var_sel]], colour = pitch_type)) + geom_point() + theme_minimal() + labs(title = glue::glue("Scatterplot of spin rate against",  input$y_var_sel, " in ", input$year_sel))})


output$spinhist <- renderPlot({ggplot(data = pitching_cleaner, aes(x = spin_rate)) + geom_histogram() + facet_wrap(~year) + theme_minimal() + labs(title = "Histogram of spin rate facted by year")})

  attendance_reactive <- reactive({
    attendance1 <- attendance_smaller |> filter(year == input$year_sel_attendance) |>  mutate(Tm = fct_reorder(Tm, .data[[input$x_sel]]))
  })
  
   attendancereactive <- reactive({
   team1 <- attendance_smaller |> filter(Tm %in% input$tm_sel)})
output$barplot <- renderPlot({ggplot(attendance_reactive(), aes(x = Tm, y = .data[[input$x_sel]])) + geom_col() + coord_flip() + theme_minimal() + labs(title = glue::glue("Barplot of ", input$x_sel, " in ", input$year_sel_attendance)) })

output$attendance_line <- renderPlot({ggplot(attendancereactive(), aes(x = year, y = `Attend/G`, colour = Tm)) + geom_line() + theme_minimal() + labs(title = "Average Attendance per Game Over Time ") })

output$audience_bar <- renderPlot({ggplot(data = tv_views_smaller, aes(x = year, y = average_audience)) + geom_col() + theme_minimal() + labs(title = "Average TV Audience per World Series Game In Each Year ") })


output$attendance_raw_data <- renderDataTable({attendance_smaller})

output$audience_raw_data <- renderDataTable({tv_views_smaller})

output$pitching_raw_data <- renderDataTable({pitching_cleaner})
}

shinyApp(ui, server)
```

# Findings

When playing around with the app, I found that there is some pretty clear difference in the hitting statistics before 2021 as opposed to after. Things like strike out percentage were higher prior to 2021 as opposed to things like opposing home runs which went up after 2021. The impact of the sticky substances is most evident in the faceted histogram of spin rate, it is quite clear that the spin rate was higher previous to 2021. In the attendance graphs as well, there is a considerable increase in both the average and total attendance following 2021. This change is somewhat less evident in the TV viewership bar plot however. What is most evident in all graphs is the 2020 season, this is especially interesting in the line graph which shows a gap during the 2020 season which is because attendance was not allowed during the COVID-19 pandemic. Additionally, the 2020 season was shortened so all statistics for that season will be offset. Another season in the data set that will be strange is the 2024 season since that one had only just begun at the time of me making the app. 

# Conclusion

If I had more time, I would have liked to add more of a modelling aspect either to the app itself or just to the project in general. It could have been interesting to assess how well spin rate predicts each of the response variables. With modelling, I could have made reactive models with liens on the scatter plot, which could have made the visualization easier to interpret. I also think looking at the correlation between spin rate and each of the response variables could have provided additional statistical information of interest.

One of the main limitations I had in making my app was that I wanted to use both the spin rate and pitch speed but I had trouble pivoting the data so that it will work. After awhile of struggling to work with the data with the pitch speed, I decided against keeping it since I realized that pitch speed is actually a dependent variable when it comes to spin rate and thus was not crucial to include it in the visualization. Another issue I had was that I could not find a data set with the MLB TV viewership data for the entire season, just the World Series. While the World Series data is still interesting, I feel that it would have been better to have a bigger TV viewership data set. Additionally, I had hoped to combine the TV and attendance data sets into one so that the comparisons could be made more clear, but due to the way they were set up this did not work. That being said, I feel that i was able to accommodate to these difficulties relatively well in an organized manner. 


